Transfer of Learned Opponent Models in Zero Sum Games

### 2021-09-20 Report/paper_draft_2021.tex
a few final issues to resolve

Iterative reasoning strategies can explain non-equilibrium play in a range of games, such as the \(p\)-beauty contest game (Nagel, 1995), dominance solvable games (C. F. Camerer, Ho, \& Chong, 2004), and various standard normal form games (Costa-Gomes, Crawford, \& Broseta, 2001). In the present study, we endow our computer opponents with limited ability for iterative reasoning and assess whether (1) human players adapt their strategy to exploit this limited reasoning of their opponent, and (2) whether they are able to generalize a learned opponent model to other games. In two experiments, participants face the same opponent (Experiment 1) or the same two opponents (Experiment 2) in three consecutive games: the well-known Rock-Paper-Scissors game, the structurally similar Fire-Water-Grass game, and a less similar Numbers (Experiment 1) or Shootout (Experiment 2) game. To foreshadow our results, we find evidence that participants transfer the learned strategy of their opponent to other games, but that this transfer is moderated by the sophistication of their opponent and the similarity of the games. Moreover, using computational modelling, we find evidence that participants switch from relying on an opponent model in the early stages of the games, to a more habitual and cognitively less demanding strategy in the later stages of the games.

@@

On average, participants obtained the lowest score in the RPS game (\(M = 0.289\), \(SD = 0.348\)), followed by Numbers (\(M = 0.31\), \(SD = 0.347\)). Participants' performance was highest in the FWG game (\(M = 0.454\), \(SD = 0.354\)). Scores in each game were significantly different from 0, the expected score of uniformly random play (RPS: \(t(51) = 7.26\), \(p < .001\); FWG: \(t(51) = 10.04\) , \(p < .001\); Numbers: \(t(51) = 7.17\), \(p < .001\)). As uniformly random play is the Nash equilibrium, this indicates successful deviation from a Nash-optimal strategy.

To assess learning within and between games, we used a 2 (condition: Level 1, Level 2) by 3 (game: RPS, FWG, Numbers) by 2 (block: first half, second half) repeated-measures ANOVA, with the first factor varying between participants. This showed a main effect of Game (\(F(2,100) = 8.54\), \(\eta^{2} = 0.05\), \(p < .001\)), indicating that average scores varied significantly over the games. Post-hoc pairwise comparisons showed that performance in the FWG game was significantly higher than in the RPS game (\(t(100) =3.78\), \(p < .001\)) and the Numbers game (\(t(100) = 3.32\) , \(p = .002\)). The score in RPS was not significantly different from the score in Numbers (\(t(100) = 0.45\) , \(p = .65\)). The main effect of Block (\(F(1,50) = 22.51\) , \(\eta^{2} = 0.03\), \(p < .001\)) shows that the score in the first half of each game (\(M = 0.29\)) was significantly lower than in the second half (\(M = 0.40\)), which indicates within-game learning. The main effect of Condition (\(F(1,50) = 5.44\), \(\eta^{2} = 0.05\), \(p = .024\)) indicates that scores were higher against the level-1 player (\(M = 0.43\)) than against the level-2 player (\(M = 0.27\)). Thus, it appears that is was harder for participants to exploit the strategy of the more sophisticated level-2 opponent than the comparatively less sophisticated level-1 opponent.

@@

Next, we explore whether learning transfer is moderated by the type of opponent and game similarity. We expected better transfer between more similar games (i.e.~better transfer from RPS to FWG than from RPS/FWG to Numbers), and worse transfer for the more sophisticated level-2 agent. Figure \ref{fig:exp1-early-score-by-opp} indicates that the pattern over the games is indeed dissimilar between level-1 and level-2 opponents. To explore this, we used a 2 (Condition: Level 1, Level 2) by 3 (Game: RPS, FWG, Numbers) repeated measures ANOVA with the first factor varying between participants. There was a main effect of Game (\(F(1.84,91.77) = 3.35\), \(\eta^{2} = 0.04\), \(p = .043\), Greenhouse-Geisser correction applied to the degrees of freedom to correct for non-sphericity). \textbf{MS: was this the only significant effect? If so, people might wonder why we next run analyses for each combination of player and game} We then ran statistical tests on early round scores by game and opponent, against the null hypothesis of 0 (no transfer). For level-1 facing players, there is evidence of learning transfer from RPS to both FWG (\(t(150) = 3.96\), \(p < .001\)) and Numbers (\(t(150) = 3.74\), \(p < .001\)). For level-2 facing players, there is evidence for transfer from RPS to the similar game FWG, albeit scores are lower than for level-1 player (\(t(150) = 2.48\), \(p = .01\)) but not to the dissimilar game of Numbers.

@@

To explore whether learning occurred within and between games, we performed a two (Condition: level-1 first, level-2 first) by two (Opponent type: level-1 or level-2) by three (Game: RPS, FWG, Shootout) by two (Encounter: first or second) repeated-measures ANOVA, with the first factor varying between participants. This showed a main effect of Game (\(F(1.85,88.7) = 11.81\), \(\eta^{2} = 0.04\), \(p < .001\)). Post-hoc pairwise comparisons between games (\(p\)-values adjusted using Holm method for multiple comparisons) indicate performance in the games increased steadily throughout the experiment, with FWG performance significantly higher than RPS (\(t(96) =2.53\), \(p = .025\)), and performance in the Shootout game significantly higher than in FWG (\(t(96) = 2.32\), \(p = .025\)). Whilst the ANOVA shows no main effect of Condition, Opponent type, or Encounter, there was a significant interaction between Game and Opponent type (\(F(1.7, 81.82) = 5.31\) ,\(\eta^{2} = 0.02\), \(p = .01\)). Follow-up analysis shows that when facing level-1 agents, scores increased steadily after each game, with FWG score significantly higher than RPS (\(t(191) = 2.70\), \(p = .03\)) and Shootout scores in turn significantly higher than FWG (\(t(191) = 3.05\), \(p = .01\)). There was no significant difference between average scores on any two games when facing level-2 agents however.

@@

To measure transfer between games, we again focus on participants' scores in the initial 5 rounds after the first round (rounds 2-6) of each game and only during the first time they interact with an opponent in a game (see Figure \ref{fig:exp2-early-score-by-opp}). We performed a two (Condition: level-1 first, level-2 first) by two (Opponent type: level-1 or level-2) by three (Game: RPS, FWG, Shootout) repeated-measures ANOVA, with the first factor varying between participants. This showed a main effect of Game (\(F(1.94,92.98) = 7.67\), \(\eta^{2} = 0.04\), \(p < .001\)). \textbf{MS: Again, if there is no interaction, there might not be justification for running tests per opponent and game }
We then ran statistical tests on early round scores by game and opponent against the null hypothesis of 0 (no transfer). For level-1 facing players, there is evidence of learning transfer from RPS to both FWG (\(t(271) = 4.99\), \(p < .001\)) and Shootout (\(t(271) = 6.66\), \(p < .001\)). For level-2 facing players, there is evidence for transfer from RPS to the similar game FWG, albeit scores are lower than for level-1 player (\(t(271) = 4.41\), \(p < .001\)). Unlike in Experiment 1, there is also evidence of transfer to the dissimilar Shootout game (\(t(271) = 3.22\), \(p = .004\)). The early round scores in the similar FWG games for level-1 and level-2 facing players are not significantly different from each other. However, the score of the players facing the level-1 opponent in Shootout is higher than that of players facing level-2 opponents ( \(t(144) = 2.45\), \(p = 0.01\)).

@@

The results of Experiment 2 confirm the earlier results of Experiment 1 in a situation where participants need to learn about two distinct opponents. As in Experiment 1, participants adapted their strategies to both level-1 and level-2 opponents to exploit deviations from Nash-optimal play. Knowledge about both types of opponents were transferred to both the similar and dissimilar game. However, transfer to the dissimilar game was easier when facing the less sophisticated agent.

@@

The self-tuning Experience Weighted Attraction (EWA) model (Teck H. Ho, Camerer, \& Chong, 2004) combines two seemingly different approaches, namely reinforcement learning and belief learning. Belief learning models are based on the assumption that players keep track of their opponent's frequency of past actions, and best respond to that. By contrast, reinforcement learning does not explicitly take into account beliefs about other players, but simply increases the probability of repeating a more rewarding action. The self-tuning EWA model has been shown to perform better than either RL or belief learning alone in various repeated games and has the advantage of having only one free parameter, the inverse temperature of the softmax choice function. The EWA model is based on updating ``Attractions'\,' for each action over time given a particular state. The attraction of action \(a\) at time \(t\) given state \(s\) is denoted as \(A_{t}(a, s)\) and updated as
\[ A_{t+1}(a,s) =  \frac{\phi \ N(t) \ A_{t}(a,s) + [ \delta + (1-\delta) \ I(a_t = a )] \ R(a,o_t) } {\phi \ N(t) + 1} \]
where \(I(x)\) is an indicator function which takes the value 1 is its argument is true and 0 otherwise, and \(R(a,o_t)\) is the reward that would be obtained from playing action \(a\) against opponent action \(o_t\). \(R(a,o_t)\) equals the actual obtained reward when \(a = a_t\), and otherwise is a counterfactual reward that would have been obtained if a different action were taken. Unlike reinforcement learning, this uses knowledge of the rules of the game to allow reinforcing actions that were not actually taken by the rewards they would have provided. The parameter \(\delta\) reflects the weight given to such counterfactual rewards. Setting \(\delta = 0\) leads to reinforcement only of actions taken, while positive values of \(\delta\) makes the update rule take into account foregone payoffs, which is similar to weighted fictitious play (Cheung \& Friedman, 1994). \(N(t)\) represents an experience weight and can be interpreted as the number of ``observation-equivalents'' of past experience. We initialise it to 1 so initial attractions and reinforcement from payoffs are weighted equally. \textbf{MS: need to better describe what \(\phi\) is, and how \(N(t)\) is determined.} In the earlier version of the EWA model (C. Camerer, Ho, \& Others, 1997), \(\phi\) and \(\delta\) were free parameters. In the self-tuning EWA model (Teck H. Ho, Camerer, \& Chong, 2004), the values of \(\delta\) and \(\phi\) are learnt from experience. For details on how this is done, we refer the reader to (Teck H. Ho, Camerer, \& Chong, 2004). \textbf{MS: Should describe how the parameters are self-tuned with equations. IG: There are many equations for each of the parameters update, phi, delta and N\ldots I'm concerned it would make for cumbersome reading, especially since EWA doesn't really fit well anyway. I've added that details of updates can be found in original camerer paper. MS: If it's three additional equations, that's not a big deal really. }

As in the models above, actions are chosen based on a softmax decision rule:
\[P_t(a,s) = \frac{\exp \{\lambda  A_{t}(a,s) \} }{\sum_{a' \in \mathcal{A}_t} \exp \{ \lambda A_{t}(a',s) \} }\]
The self-tuning EWA has one free parameter: the inverse temperature of the softmax decision rule (\(\lambda\)).

@@

For both experiments, we fitted all models to individual participant data by maximum likelihood estimation. \textbf{MS: need to mention the use of e.g.~DEoptim} We use the Bayesian Information Criterion (BIC) to determine the best fitting model for each participant.

For Experiment 1, we fitted a total of 5 models: a baseline model assuming random play (Nash), the Bayesian Cognitive Hierarchy model allowing transfer between games (BCH\_BT) and without transfer between games (BCH\_NBT), as well as a reinforcement learning (RL), and finally a self-tuning EWA model with the same state space (EWA).

@@

\textbf{MS: Please present the models in the figures in the order mentioned in text (i.e., Nash, RL, EWA, BCH), and when providing versions of each, do so in the same order as well, i.e.~x\_NT, x\_WT, x\_BT}

Figure~\ref{fig:exp1-comp-models} shows the results for Experiment 1. We can see that the RL model clearly described most participants' behaviour best, followed by the random (Nash) model. Only a few participants were best described by one of the BCH models or the EWA model. Looking at BIC weights (Figure ~\ref{fig:xp1-BIC-weigths}), we confirm this. RL models have high BIC weights when they best fit the participants, and very few instances have high BIC weights for models other then RL, which fits the picture drawn by the histogram. \textbf{MS: The BIC weight don't add much information, so maybe we could leave them out }

@@

{\centering \includegraphics{paper_draft_2021_files/figure-latex/xp1-BIC-weigths-1}

\caption{Model BIC weights for participants in Experiment 1.}\label{fig:xp1-BIC-weigths}
\end{figure}

In Experiment 2, we can see from Figure~\ref{fig:exp2-comp-models} that the RL model was again more successful than the Bayesian Cognitive Hierarchy models (with or without transfer) and the EWA model in fitting participants' action choices. This is also reflected in BIC weights in Figure \ref{fig:xp2-BIC-weigths}.

@@

\caption{Experiment 2 Histogram of best fitting computational models by condition}\label{fig:exp2-comp-models}
\end{figure}

{\centering \includegraphics[width=\textwidth]{paper_draft_2021_files/figure-latex/xp2-BIC-weigths-1}

\caption{BIC weights for each model and participant in Experiment 2.}\label{fig:xp2-BIC-weigths}
\end{figure}

@@

\caption{Experiment 1 Likelihood by trial, game and opponent faced}\label{fig:exp1-lik-by-tr}
\end{figure}

@@

The switching model (AIC = 16,879.97, BIC = 16,936.81) in Experiment 2 again fits better than the restricted non-switching model (AIC = 16,938.41, BIC = 16,952.62), \(\chi^2(6) = 70.44\), \(p < .001\). The posterior probabilities of the strategies (Figure~\ref{fig:exp2-posteriors-plot}) show very clear evidence of strategy switching across games and stages, from using a BCH model in the initial rounds to an RL strategy later on.

@@

The results of our first experiment show that the majority of participants learn to adapt to the opponent strategy over multiple interactions and generalise this learning to a similar game. Performance in early rounds indicated that transfer to the more dissimilar game was moderated by the degree of sophistication of the opponent, with evidence for transfer when players face the less sophisticated agent but not the more sophisticated one. In the second experiment, participants faced both types of opponents, which allows for a stronger test of opponent modelling, as participants would need to learn a different strategy for each opponent within a game. In Experiment 1, participants could learn a single strategy for each game, making opponent modelling possibly less pertinent. Experiment 2 also offers more opportunities to test transfer of a learnt opponent model. The results on learning transfer confirmed the findings from the first experiment. Again, there was clear evidence of transfer in the early rounds of the later games. We also found that learning transfer is moderated by the type of opponent faced: Evidence of transfer was weaker for the level-2 opponent as compared to the level-1 opponent. That transfer was less evident for the more sophisticated level-2 opponent in both experiments may be due to a higher difficulty of learning that opponent's strategy. If it is more difficult to establish a model of the level-2 opponent, there is likely less knowledge to transfer to the new game. A player cannot transfer what they have not learnt.

What exactly did the players learn in RPS that allowed them to beat the opponent in FWG and Shootout? what did the players learn specifically about their opponent's strategy and what form did this learning take?

One possible answer is that participants learned simple rules based on last round play. For instance, ``play scissors whenever my opponent played rock in last round,'' or ``play paper whenever the last round play was either rock or scissors.'' These are the type of strategies that are learned by the model-free reinforcement learning we used in our computational modelling. While this strategy fitted participants' actions the best overall, there are at least two reasons why this account is not satisfactory as a complete description of participants' behaviour. Firstly, the learned strategies are not transferable to new games. There is no simple way to map ``play scissors whenever my opponent played rock in last round'' in the RPS game to ``play grass whenever my opponent played fire in last round.'' Such a mapping may be possible by translating the rules and structure from RPS to FWG, but model-free reinforcement learning lacks the tools to do this quickly enough: Model-free reinforcement learning would need to start from scratch in each new game, yet we found evidence that participants could successfully exploit their opponent's strategy in early rounds of new games. Secondly, a reinforcement learning strategy would fare equally well against the level-1 and level-2 opponent. Whilst choosing different actions, the contingency between the state (last round play) and actions is the same for both opponents. Yet, we found that participants performed better against the level-1 opponent compared to the level-2 opponent. The difference in performance between the two types of opponent indicate that the actions of the more sophisticated level-2 opponent, or the best response to these, were somehow more difficult to predict.

We are left with two possible explanations: First, it is possible that participants discovered a heuristic that allowed them to beat their opponent without explicitly modelling their strategy, and that this heuristic is transferable to new games. Because of the cyclicity in action choices (e.g., rock beats scissors, scissors beats paper, paper beats rock), it is possible to beat a level-2 opponent most of the time by following a simple rule: Play in the next round whatever the opponent played in the last round. This is a rule that wins and is transferable to other games as it does not depend on action labels. In the same vein, a heuristic that beats a level-1 player can be stated as ``Choose the action that would have been beaten by my previous action.'' Intuitively, it seems that the heuristic for the level-2 player is simpler than that for the level-1 player, which is at odds with the difference in performance.

@@

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

@@

@@

@@

\endgroup

>>>>

Iterative reasoning strategies can explain non-equilibrium play in a range of games, such as the \(p\)-beauty contest game (Nagel, 1995), dominance solvable games (C. F. Camerer, Ho, \& Chong, 2004), and various standard normal form games (Costa-Gomes, Crawford, \& Broseta, 2001). In the present study, we endow our computer opponents with limited ability for iterative reasoning and assess whether (1) human players adapt their strategy to exploit this limited reasoning of their opponent, and (2) whether they are able to generalize a learned opponent model to other games. In two experiments, participants face the same opponent (Experiment 1) or the same two opponents (Experiment 2) in three consecutive games: the well-known Rock-Paper-Scissors game, the structurally similar Fire-Water-Grass game, and a less similar Numbers (Experiment 1) or Shootout (Experiment 2) game. To foreshadow our results, we find evidence that participants transfer the learned strategy of their opponent to other games, and that this transfer is moderated by the sophistication of their opponent and the similarity between the games. Moreover, using computational modelling, we find evidence that participants switch from relying on an opponent model in the early stages of the games, to a more habitual and cognitively less demanding strategy in the later stages of the games.

@@

First, we focus on participants scores in each game (see Figure \ref{fig:exp1-avg-scores-game}). On average, participants obtained the lowest score in the RPS game (\(M = 0.289\), \(SD = 0.348\)), followed by Numbers (\(M = 0.31\), \(SD = 0.347\)). Participants' performance was highest in the FWG game (\(M = 0.454\), \(SD = 0.354\)). Scores in each game were significantly different from 0, the expected score of uniformly random play (RPS: \(t(51) = 7.26\), \(p < .001\); FWG: \(t(51) = 10.04\) , \(p < .001\); Numbers: \(t(51) = 7.17\), \(p < .001\)). As uniformly random play is the Nash equilibrium, this indicates successful deviation from a Nash-optimal strategy.

To assess learning within and between games, we estimate a linear mixed-effects model with score as the dependent variable, and Condition (Level 1 or Level 2), Game (RPS, FWG, Numbers) and Block (first-half , second-half) as independent variables, including all main effects and interactions. The model included random intercepts for participants. The results show a main effect of Game (\(F(2,250) = 12.90\), \(\eta^{2} = 0.09\), \(p < .001\)) indicating that performance differs between the games. Post-hoc pairwise comparisons show that performance in the FWG game was significantly higher than in the RPS game (\(t(100) =3.78\), \(p < .001\)) and the Numbers game (\(t(100) = 3.32\) , \(p = .002\)). The score in RPS was not significantly different from the score in Numbers (\(t(100) = 0.45\) , \(p = .65\)). The main effect of Block (\(F(2,250) = 15.16\) , \(\eta^{2} = 0.06\), \(p < .001\)) shows that the score in the first half of each game (\(M = 0.29\)) was significantly lower than in the second half (\(M = 0.40\)), which indicates within-game learning. The main effect of Condition (\(F(1,50) = 5.44\), \(\eta^{2} = 0.10\), \(p = .024\)) indicates that scores were higher against the level-1 player (\(M = 0.43\)) than against the level-2 player (\(M = 0.27\)). Thus, it appears that is was harder for participants to exploit the strategy of the more sophisticated level-2 opponent than the comparatively less sophisticated level-1 opponent. There was also an interaction between Game and Condition (\(F(2,250) = 4.52\) , \(\eta^{2} = 0.03\), \(p = .01\)). \textbf{MS: What does this interaction indicate?}

@@

Next, we explore whether learning transfer is moderated by the type of opponent and game similarity. We expected better transfer between more similar games (i.e.~better transfer from RPS to FWG than from RPS/FWG to Numbers), and worse transfer for the more sophisticated level-2 agent. Figure \ref{fig:exp1-early-score-by-opp} indicates that the pattern over the games is indeed dissimilar between level-1 and level-2 opponents. We estimates a linear mixed-effects model with the early round score as the dependent variable, and Condition (Level 1 or Level 2) and Game (RPS, FWG, Numbers), as well as their interaction, as independent variables. The model included random intercepts for participants. The results indicated a main effect of Game (\(F(2,100) = 3.35\), \(p = .04\)), but no effect for condition. Further investigation of the effects of game on early score using planned contrasts revealed that early scores in the FWG games are significantly higher than in the RPS ( \(t(99.99) = -2.32\), \(p = 0.02\)).

We next ran statistical tests on early round scores by game and opponent, against the null hypothesis of 0 (no transfer). For level-1 facing players, there is evidence of learning transfer from RPS to both FWG (\(t(150) = 3.96\), \(p < .001\)) and Numbers (\(t(150) = 3.74\), \(p < .001\)). For level-2 facing players, there is evidence for transfer from RPS to the similar game FWG, albeit scores are lower than for level-1 player (\(t(150) = 2.48\), \(p = .01\)) but not to the dissimilar game of Numbers.

@@

To explore whether learning occurred within and between games, we estimated a linear mixed-effects model, with score as the dependent variable and Condition (level-1 first, level-2 first), Opponent type (level-1 or level-2), Game (RPS, FWG, Shootout) and Encounter (first or second) as the independent variables, including all main effects and interactions. The model included random intercepts for participants. We found a main effect of Game (\(F(2,528) = 15.32\), \(\eta^{2} = 0.04\), \(p < .001\)). Post-hoc pairwise comparisons between games (\(p\)-values adjusted using Holm method for multiple comparisons) indicate performance in the games increased steadily throughout the experiment, with FWG performance significantly higher than RPS (\(t(96) =2.53\), \(p = .025\)), and performance in the Shootout game significantly higher than in FWG (\(t(96) = 2.32\), \(p = .025\)). Whilst the model shows no main effect of Condition, Opponent type, or Encounter, there was a significant interaction between Game and Opponent type (\(F(2, 528) = 5.78\) ,\(\eta^{2} = 0.02\), \(p = .003\)). Follow-up analysis differentiating performance by type of opponent faced shows that when facing level-1 agents, scores increased steadily after each game, with FWG score significantly higher than RPS (\(t(191) = 2.70\), \(p = .03\)) and Shootout scores in turn significantly higher than FWG (\(t(191) = 3.05\), \(p = .01\)). There was no significant difference between scores on games when facing level-2 agents however.

@@

To measure transfer between games, we again focus on participants' scores in the initial 5 rounds after the first round (rounds 2-6) of each game and only during the first time they interact with an opponent in a game (see Figure \ref{fig:exp2-early-score-by-opp}). We estimated a linear mixed-effects model with the early round scores as the dependent variable, and Condition (level-1 first, level-2 first), Opponent type (level-1 or level-2) and Game (RPS, FWG, Shootout) as independent variables, including all main effects and interactions. The model also included random intercepts for participants. We found a main effect of Game (\(F(2,240) = 6.57\), \(p = .002\)) and an interaction effect between Game and Opponent type (\(F(2,240) = 3.34\), \(p = .037\)). \textbf{MS: What do these effects indicate? Report post-hoc comparisons between games, and e.g.~differences between opponents for each game, here. The later tests against 0 are not adequate to understand the effects here.}

We then ran statistical tests on early round scores by game and opponent against the null hypothesis of 0 (no transfer). For level-1 facing players, there is evidence of learning transfer from RPS to both FWG (\(t(271) = 4.99\), \(p < .001\)) and Shootout (\(t(271) = 6.66\), \(p < .001\)). For level-2 facing players, there is evidence for transfer from RPS to the similar game FWG, albeit scores are lower than for level-1 player (\(t(271) = 4.41\), \(p < .001\)). Unlike in Experiment 1, there is also evidence of transfer to the dissimilar Shootout game (\(t(271) = 3.22\), \(p = .004\)). The early round scores in the similar FWG games for level-1 and level-2 facing players are not significantly different from each other. However, the score of the players facing the level-1 opponent in Shootout is higher than that of players facing level-2 opponents (\(t(144) = 2.45\), \(p = 0.01\)).

@@

The results of Experiment 2 confirm earlier findings from Experiment 1 on learning transfer in a situation where participants need to learn about two distinct opponents. As in Experiment 1, participants adapted their strategies to both level-1 and level-2 opponents to exploit deviations from Nash-optimal play. Knowledge about both types of opponents were transferred to both the similar and dissimilar game. This transfer was moderated by the degree of similarity between games, as well as opponent type. The interaction effect showed that transfer to the dissimilar game was easier when facing the less sophisticated agent.

@@

The self-tuning Experience Weighted Attraction (EWA) model (Teck H. Ho, Camerer, \& Chong, 2004) combines two seemingly different approaches, namely reinforcement learning and belief learning. Belief learning models are based on the assumption that players keep track of their opponent's frequency of past actions, and best respond to that. By contrast, reinforcement learning does not explicitly take into account beliefs about other players, but simply increases the probability of repeating a more rewarding action. The self-tuning EWA model has been shown to perform better than either RL or belief learning alone in various repeated games and has the advantage of having only one free parameter, the inverse temperature of the softmax choice function. The EWA model is based on updating ``Attractions'\,' for each action over time given a particular state. The attraction of action \(a\) at time \(t\) given state \(s\) is denoted as \(A_{t}(a, s)\) and updated as
\[ A_{t+1}(a,s) =  \frac{\phi(t) \ N(t) \ A_{t}(a,s) + [ \delta_{a}(t) + (1-\delta_{a}(t)) \ I(a_t = a )] \ R(a,o_t) } {\phi(t)N(t) + 1} \]
where \(I(x)\) is an indicator function which takes the value 1 is its argument is true and 0 otherwise, and \(R(a,o_t)\) is the reward that would be obtained from playing action \(a\) against opponent action \(o_t\). \(R(a,o_t)\) equals the actual obtained reward when \(a = a_t\), and otherwise is a counterfactual reward that would have been obtained if a different action were taken. Unlike reinforcement learning, this uses knowledge of the rules of the game to allow reinforcing actions that were not actually taken by the rewards they would have provided. The parameter \(\delta\) reflects the weight given to such counterfactual rewards. Setting \(\delta = 0\) leads to reinforcement only of actions taken, while positive values of \(\delta\) makes the update rule take into account foregone payoffs, which is similar to weighted fictitious play (Cheung \& Friedman, 1994). \(N(t)\) represents an experience weight and can be interpreted as the number of ``observation-equivalents'' of past experience. We initialise it to 1 so initial attractions and reinforcement from payoffs are weighted equally.

In the earlier version of the EWA model (C. Camerer, Ho, \& Others, 1997), \(\phi\) and \(\delta\) were time-invariant free parameters. In the self-tuning EWA model (Teck H. Ho, Camerer, \& Chong, 2004), the values of \(\delta\) and \(\phi\) are learnt from experience. Over time, \(\delta\) is updated as
\[\delta_{a}(t) = \begin{cases} 1 & \text{if }  R(a,o_{t}) \geq r_{t}  \\
0 & \text{otherwise} \end{cases}\]
The \(\phi_{t}\) parameter can be interpreted as a discount of prior experience, modelling either limited agent memory or changes in the game conditions. At the core, \(\phi(t)\) depends on a surprise index \(S_{p}(t)\):
\[\phi(t) = 1 - \frac{1}{2}S_{p}(t) ,\]
where \(S_{p}(t)\) quantifies how the opponent deviates from past play. It is calculated in turn through the cumulative history of play \(h_{k}(t)\):
\[h^{k}(t)= \frac{\sum_{\tau =1}^t I(o^k,o_{\tau})}{t}, \]
as well as an indicator vector of the most recent opponent play:
\[r^k(t) = I(o^k,o_{t}), \]
Where \(I\) is the indicator function as defined above. To get the surprise index, we simply sum all the squared deviations between the cumulative history vector \(h^{k}(t)\) and the immediate history \(r^k(t)\):
\[S_{p}(t) = \sum_{k=1}^{|\mathcal{A}_g|} (h^{k}(t) - r^k(t))^2 \]
\textbf{MS: why is there a superscript in \(h^k\) and \(r^k\)? And is the indicator variable \(I(o^k = o_{t})\)? And is \(\delta_a(t)\) always 0 or 1?}

For more details on the self-tuning EWA model, we refer the reader to (Teck H. Ho, Camerer, \& Chong, 2004). As in the RL model above, actions are chosen based on a softmax decision rule:

\[P_t(a,s) = \frac{\exp \{\lambda  A_{t}(a,s) \} }{\sum_{a' \in \mathcal{A}_t} \exp \{ \lambda A_{t}(a',s) \} } \]
The self-tuning EWA has one free parameter: the inverse temperature of the softmax decision rule (\(\lambda\)).

@@

For both experiments, we fitted all models to individual participant data by maximum likelihood estimation using the DEoptim R package (Mullen, Ardia, Gil, Windover, \& Cline, 2011). We use the Bayesian Information Criterion (BIC) to determine the best fitting model for each participant.

For Experiment 1, we fitted a total of 5 models: a baseline model assuming random play (Nash), the Bayesian Cognitive Hierarchy model allowing transfer between games (BCH\_BT) and without transfer between games (BCH\_NT), as well as a reinforcement learning (RL), and finally a self-tuning EWA model with the same state space (EWA).

@@

Figure~\ref{fig:exp1-comp-models} shows the results for Experiment 1. We can see that the RL model clearly described most participants' behaviour best, followed by the random (Nash) model. Only a few participants were best described by one of the BCH models or the EWA model.

@@

In Experiment 2, we can see from Figure~\ref{fig:exp2-comp-models} that the RL model was again more successful than the Bayesian Cognitive Hierarchy models (with or without transfer) and the EWA model in fitting participants' action choices.

@@

\caption{Experiment 2 histogram of best fitting computational models by condition.}\label{fig:exp2-comp-models}
\end{figure}

@@

\caption{Experiment 1 likelihood by trial, game and opponent faced}\label{fig:exp1-lik-by-tr}
\end{figure}

@@

In Experiment 2, the switching model (AIC = 16,879.97, BIC = 16,936.81) again fits better than the restricted non-switching model (AIC = 16,938.41, BIC = 16,952.62), \(\chi^2(6) = 70.44\), \(p < .001\). The posterior probabilities of the strategies (Figure~\ref{fig:exp2-posteriors-plot}) show very clear evidence of strategy switching across games and stages, from using a BCH model in the initial rounds to an RL strategy later on.

@@

The results of our first experiment show that the majority of participants learnt to adapt to their opponent's strategy over multiple interactions, and generalised this learning to a similar game. Performance in early rounds indicated that learning transfer was moderated by the degree of similarity between games. In the second experiment, participants faced both types of opponents, allowing for a stronger test of opponent modelling, as participants would need to learn a different strategy for each opponent. In Experiment 1, participants could learn a single strategy for each game, making opponent modelling possibly less pertinent. In Experiment 2, we again found clear evidence of transfer in the early rounds of the later games. In addition to the moderation of transfer by the similarity between games, we also found that learning transfer was moderated by the type of opponent faced: Transfer was harder when facing the level-2 opponent as compared to the level-1 opponent. That evidence for transfer was weaker when facing the more sophisticated level-2 opponent may be due to a higher difficulty of learning that opponent's strategy. If it is more difficult to establish a model of the level-2 opponent, there is likely less knowledge to transfer to the new game. A player cannot transfer what they have not learnt.

What exactly did the players learn in the first game (RPS) that allowed them to beat their opponent in the later games (FWG, and Numbers or Shootout)? what did the players learn specifically about their opponent's strategy and what form did this learning take?
One possible answer is that participants learned simple rules based on last round play. For instance, ``play scissors whenever my opponent played rock in last round,'' or ``play paper whenever the last round play was either rock or scissors.'' These are the type of strategies that are learned by the model-free reinforcement learning we used in our computational modelling. While this strategy fitted participants' actions the best overall, there are at least two reasons why this account is not satisfactory as a complete description of participants' behaviour. Firstly, the learned strategies are not transferable to new games. There is no simple way to map ``play scissors whenever my opponent played rock in last round'' in the RPS game to ``play grass whenever my opponent played fire in last round.'' Such a mapping may be possible by translating the rules and structure from RPS to FWG, but model-free reinforcement learning lacks the tools to do this: Model-free reinforcement learning would need to start from scratch in each new game, yet we found evidence that participants could successfully exploit their opponent's strategy in early rounds of new games. Secondly, a reinforcement learning strategy would fare equally well against the level-1 and level-2 opponent. Whilst choosing different actions, the contingency between the state (last round play) and actions is the same for both opponents. Yet, we found that participants performed better against the level-1 opponent compared to the level-2 opponent. The difference in performance between the two types of opponent indicates that the actions of the more sophisticated level-2 opponent, or the best response to these, were somehow more difficult to predict.

We are left with two possible explanations: First, it is possible that participants discovered a heuristic that allowed them to beat their opponent without explicitly modelling their strategy, and that this heuristic is transferable to new games. Because of the cyclicity in action choices (e.g., rock beats scissors, scissors beats paper, paper beats rock), it is possible to beat a level-2 opponent most of the time by following a simple rule: Play in the next round whatever the opponent played in the last round. This is a rule that wins and is transferable to other games, as it does not depend on action labels. In the same vein, a heuristic that beats a level-1 player can be stated as ``Choose the action that would have been beaten by my previous action.'' Intuitively, it seems that this heuristic for a level-2 player is simpler than that for a level-1 player, which is at odds with the finding that participants performed better against the level-1 opponent.

@@

@@

\leavevmode\hypertarget{ref-mandler2004foundations}{}%
Mandler, J. M. (2004). \emph{The foundations of mind: Origins of conceptual thought}. Oxford University Press.

@@

\leavevmode\hypertarget{ref-R-DEoptim}{}%
Mullen, K., Ardia, D., Gil, D., Windover, D., \& Cline, J. (2011). {DEoptim}: An {R} package for global optimization by {D}ifferential {E}volution. \emph{Journal of Statistical Software}, \emph{40}(6), 1--26. \url{https://doi.org/10.18637/jss.v040.i06}

@@
